Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-07-28 resubmission/blinded-paper-resubmit.tex
Update varios ajustes

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

  \author{Juan Carlos Castillo}
            \affil{%
                  Departamento de Sociología, Universidad de Chile
              }
          \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
          \affil{%
                  Núcleo milenio de desigualdades y oportunidades
                  digitales (NUDOS)
              }
        \author{Andreas Laffert}
            \affil{%
                  Instituto de Sociología, Pontificia Universidad
                  Católica de Chile
              }
        \author{Kevin Carrasco}
            \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
        \author{Julio Iturra-Sanhueza}
            \affil{%
                  International Graduate School of Social Sciencies
                  (BIGSSS), University of Bremen, Germany
              }

\date{}
\begin{document}
\maketitle
\begin{abstract}
This study investigates the relationship between perceptions of economic
inequality, meritocratic beliefs, and preferences for market justice in
Chile between 2016 and 2023. Using six waves of panel data from the
Chilean Longitudinal Social Survey - ELSOC (\(N_{observations}\) =
8,643; \(N_{individuals}\) = 1,687), the analysis examines how
subjective assessments of inequality shape attitudes toward the role of
merit in access to key social services such as healthcare, education,
and pensions. Results indicate that rising perceptions of inequality are
associated with greater support for redistributive policies; however,
individuals with strong meritocratic convictions are more likely to
legitimize existing disparities. The study also considers the influence
of major social movements during this period, which appear to have
reshaped public discourse and perceptions of fairness. These findings
contribute to a deeper understanding of how beliefs about justice and
equity evolve in contexts marked by persistent inequality and entrenched
market-oriented frameworks. \newline \textbf{Keywords}: Economic
inequality, meritocracy, market justice, Chile, public preferences,
inequality perception
\end{abstract}

@@

Since 1980, economic inequality and wealth concentration have
dramatically increased worldwide, becoming one of the main challenges
for the social sciences. Globally, in 2021, less than 50\% of the world
population owned only 2\% of the wealth, while the richest 10\%
concentrated 76\%, and the wealthiest 1\% captured nearly 38\% of total
assets (\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}). This
context of economic disparity has sparked renewed interest in studying
not only the objective aspects of inequality, such as income and access
to resources, but also its subjective dimensions, including perceptions,
beliefs, and associated attitudes
(\citeproc{ref-janmaat_subjective_2013}{Janmaat, 2013}). This has led,
among other reasons, to widespread concern in the social sciences about
the political delegitimization that economic inequality can produce
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}).
Understanding how people perceive inequality is crucial, as these
perceptions can influence how societies comprehend and justify the
distribution of goods and services.

The perception of economic inequality can be understood as an
individual's subjective assessment of how resources are allocated among
members of a given society (\citeproc{ref-akyelken_urban_2020}{Akyelken,
2020}). Regardless of their measurement, various studies have shown that
this perception often underestimates the gap between the rich and the
poor, which could have implications for attitudes toward the
distribution of goods and services
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-schroder_income_2017}{Schröder, 2017}). Moreover, in
recent years, there has been a discussion about the subjective and
objective aspects of inequality, demonstrating inconsistencies between
the two (\citeproc{ref-trump_income_2018}{Trump, 2018}). This tension is
relevant because advancing the study of perceptions could help
understand how perceived inequality affects the distribution of
resources within a society (Becker, 2020).

Perception of economic inequality has been associated with important
outcomes such as redistributive preferences, justification of
inequality, and legitimacy of the economic system
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_attitudes_2020}{García-Sánchez et al.,
2020}, \citeproc{ref-garcia-sanchez_perceptions_2018}{2018}). Research
shows that greater perceived inequality fosters stronger preferences for
redistribution, regardless of actual inequality levels
(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
Treisman, 2018}). However, other findings suggest that heightened
perceptions of income gaps between high- and low-paying occupations may
lead to greater justification of inequality
(\citeproc{ref-Castillo2011}{Castillo, 2011}). A less explored area in
this regard concerns market justice, which denotes the degree to which
individuals consider just that social goods and services (e.g., health,
pensions, education) are allocated according to individual contribution,
competition, and ability to pay
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lane_market_1986}{Lane, 1986};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Market justice attitudes
reflect the belief that the market promotes procedural
fairness---equality of opportunity---so that outcomes result from
individual achievement (\citeproc{ref-lane_market_1986}{Lane, 1986}).
Indeed, Lindh (\citeproc{ref-lindh_public_2015}{2015}) shows that market
justice preferences are stronger in countries with higher private
spending on services than in those with more comprehensive welfare
systems. From this perspective, perceptions of economic inequality may
be key, as lower perceived gaps can reinforce acceptance of free-market
systems and justify existing inequalities.

Meritocracy posits that inequality can be legitimized through
distributive criteria such as effort and talent
(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
\citeproc{ref-young_rise_1962}{Young, 1962}). Previous studies have
demonstrated that people with stronger meritocratic beliefs tend to
perceive less inequality as they attribute economic differences to
individual achievements (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
\citeproc{ref-wilson_role_2003}{Wilson, 2003}) and also to justify more
inequality as these beliefs are associated with attitudes that
legitimize status differences
(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In a context
like Chile, where the distribution of goods and services is
predominantly governed by market logics strongly introduced during the
military dictatorship (1973-1989)
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}), these beliefs can
play a crucial role in the acceptance of social inequalities.
Incorporating the perception of meritocracy in this literature allows
for an understanding of how attitudes toward inequality are shaped not
only by the perception of economic gaps, but also by beliefs about how
these gaps are justified. Furthermore, these beliefs are often
consolidated from an early age, reinforced by institutions promoting
values such as effort and individual skills to climb socially
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian, 2014}).

The perception of meritocracy and the perception of economic inequality
would interact intricately in shaping preferences for market justice. On
the one hand, a stronger perception of merit as a criterion for
distribution could minimize the perception of economic gaps, thus
justifying an unequal system
(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). On the other hand, the
perception of inequality could moderate the impact of meritocracy, as a
higher perception of economic gaps might challenge the idea that these
are solely based on merit.

The primary objective of this study is to analyze the interplay between
perceptions of inequality and meritocracy and their joint influence on
preferences for market justice in Chile from 2016 to 2023, using
longitudinal data from the ELSOC survey. This interaction is expected to
provide a more comprehensive explanation of attitudes toward market
justice in a country characterized by high inequality and strong
free-market influence (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
\citeproc{ref-flores_top_2020}{Flores et al., 2020}). Additionally, this
analysis seeks to elucidate how political and social
contingencies---such as the 2019 and 2022 social movements---might have
moderated these relationships by prompting more critical reflection on
the commodification of social services. Recognizing the temporal
dimension in shaping market justice preferences is essential, given that
such preferences are not static but are influenced by historical and
contextual factors that challenge or reaffirm prevailing social norms.
In this regard, variations in perceived economic inequality and
meritocracy over time can affect how individuals endorse market-based
approaches. Consequently, public opinion may shift from supporting
market justice to embracing redistributive policies aimed at mitigating
inequalities.

@@

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit. It is possible to trace the concept
to the understanding of Lane (\citeproc{ref-lane_market_1986}{1986}),
who makes a contrast between market justice and political justice. The
author defines market justice as a system of ``earned deserts'', whereby
individuals are seen as deserving of a determined distributive outcome
due to their effort and skills. In contrast, political justice
emphasizes principles of equality and need, which are often represented
by the welfare state action through social policies. An important remark
is that the principles of market justice rely on the assumption that
markets are neutral and self-regulating arenas, where individuals are
treated fairly because they face the same formal rules of engagement
(\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently, the
legitimacy of market justice stems from the assumption that inequalities
are not only inevitable but fair---so long as the rules are transparent
and opportunities are open. In this way, market justice provides a moral
justification for inequality by framing it as a necessary outcome of
individual responsibility.

@@

The study of market justice preferences has increasingly focused on how
they are shaped by individuals' socioeconomic position, normative
beliefs, and the institutional context in which they are embedded.
Across the literature, there is empirical evidence suggesting that
individuals in more advantaged socioeconomic positions---those with
higher occupational class, income, and education---are more likely to
support market justice principles (\citeproc{ref-koos_moral_2019}{Koos
\& Sachweh, 2019}; \citeproc{ref-svallfors_political_2007}{Svallfors,
2007}). This tendency reflects not only material self-interest but also
a broader moral economy, in which winners of the market system
internalize justifications for the status quo. At the same time,
political ideology also plays a role ---such as economically
conservative values --- where more right-wing individuals show higher
support for meritocracy and more skepticism towards redistribution. This
is particularly salient in countries with more restricted public
provision of social services. For example, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) in Chile and, Lee and
Stacey (\citeproc{ref-lee_fairness_2023}{2023}) in Australia show that
those with right-leaning individuals are more supportive of market-based
distribution of welfare. Beyond individual characteristics,
country-level institutions also play a central role. In liberal welfare
regimes like those of the United States or the United Kingdom, market
justice preferences are more widespread, while in coordinated or
social-democratic regimes---such as Sweden or Germany---citizens are
generally more critical of market-based inequalities
(\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
that market justice is also shaped contextually, which is represented by
the development of welfare institutions, in line with the
policy-feedbacks literature
(\citeproc{ref-campbell_policy_2012}{Campbell, 2012};
\citeproc{ref-pierson_when_1993}{Pierson, 1993}).

@@

Perceptions of inequality have been associated with attitudes about
market justice (\citeproc{ref-kluegel_social_1995a}{Kluegel et al.,
1995}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Research
indicates that lower perceived inequality can reinforce support for
market-based distributive arrangements by suggesting that the system is
fair and that outcomes reflect effort and ability
(\citeproc{ref-kuhn_eye_2011}{Kuhn, 2011}). In contrast, when inequality
is perceived as excessive or structurally determined, individuals could
question the legitimacy of market justice and become more supportive of
redistributive policies
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}).

Perceptions of inequality refer to individuals' subjective evaluations
of the extent, causes, and consequences of income and wealth
disparities. Unlike objective measures such as the Gini index, perceived
inequality captures how individuals make sense of distributive
hierarchies in their everyday lives, shaped by reference groups, social
comparisons, and information environments
(\citeproc{ref-garcia-castro_perceiving_2020}{García-Castro et al.,
2020}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
Treisman, 2018}; \citeproc{ref-mijs_stratified_2016}{Mijs, 2016}).
Scholars have proposed multiple dimensions of perceived inequality,
including its magnitude (how significant are the gaps), vertical
structure (between which groups), the trend over time (increasing or
decreasing), and legitimacy (whether it is just or not)
(\citeproc{ref-engelhardt_what_2018}{Engelhardt \& Wagener, 2018};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}). These dimensions encompass both cognitive and normative aspects
of perceptions of inequality and can vary across societies and social
groups, depending on exposure, ideology, and personal experience
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
2018}).

@@

In addition to influencing individual attitudes toward inequality,
meritocratic beliefs can contribute to social division and the
stigmatization of disadvantaged groups. Recent research has demonstrated
that exposure to meritocratic narratives can reinforce the belief that
poverty results from individual failings rather than systemic
conditions, reducing support for redistributive measures and increasing
the stigmatization of the poor (\citeproc{ref-hoyt_mindsets_2023}{Hoyt
et al., 2023}). This reinforces negative stereotypes and reduces empathy
toward individuals from lower socioeconomic backgrounds. Moreover,
Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argues
that meritocratic narratives can serve as feedback mechanisms that shape
public opinion and well-being by framing individuals' understanding of
welfare outcomes as deserved or undeserved within existing institutional
structures. This psychological mechanism highlights the normative power
of meritocracy in stabilizing unequal systems by shaping political
attitudes and personal perceptions of success and failure.

@@

\(H1\): Higher perceived economic inequality is associated with less
market justice preferences.

\(H2\): Higher meritocratic beliefs are associated with higher market
justice preferences.

>>>>

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
Several countries have experienced a shift toward the privatization and
commodification of public goods, welfare policies, and social services.
In Latin America, Chile stands out as a paradigmatic case where this
trend has led to the extensive marketization of essential services. From
a moral economy perspective, the extent to which individuals consider it
fair for access to such services to depend on market criteria has been
conceptualized as market justice preferences. This study investigates
the relationship between perceptions of economic inequality,
meritocratic beliefs, and market justice preferences in Chile between
2016 and 2023. Using six waves of panel data from the Chilean
Longitudinal Social Survey -- ELSOC (\(N_{observations}\) = 8,643;
\(N_{individuals}\) = 1,687), the analysis examines how subjective
assessments of inequality shape attitudes toward the role of merit in
access to key social services such as healthcare, education, and
pensions. The findings show that greater perceived inequality is
associated with lower market justice preferences. However, individuals
who believe that effort is rewarded are more likely to legitimize
existing disparities. In contrast, the perception that talent is
rewarded shows a negative effect on market justice preferences; an
effect that intensifies as perceived inequality increases over time. The
study also considers the influence of major social movements during this
period, which appear to have reshaped public discourse on justice and
fairness. These findings contribute to a deeper understanding of how
distributive beliefs evolve in contexts marked by persistent inequality
and entrenched neoliberal frameworks \newline \textbf{Keywords}:
Economic inequality, meritocracy, market justice, Chile, public
preferences, inequality perception
\end{abstract}

@@

Since 1980, economic inequality and wealth concentration have
dramatically increased worldwide, becoming one of the main challenges
for the social sciences. Globally, in 2021, less than 50\% of the world
population owned only 2\% of the wealth, while the richest 10\%
concentrated 76\%. Futhermore, the wealthiest 1\% captured nearly 38\%
of total assets (\citeproc{ref-chancel_world_2022}{Chancel et al.,
2022}). This context of economic disparity has sparked renewed interest
in studying not only the objective aspects of inequality, such as income
and access to resources, but also its subjective dimensions, including
perceptions, beliefs, and associated attitudes
(\citeproc{ref-janmaat_subjective_2013}{Janmaat, 2013}). Such dimensions
are deemed as key to understand the political delegitimation that
economic inequality can elicit
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}).
Therefore, understanding how people perceive inequality is crucial, as
these perceptions can influence how societies comprehend and justify (or
not) inequality in the distribution of goods and services, as well as
their capacity to funnel political discontent and eventually opposition
to the current economic distribution.

The justification of economic inequality has a long tradition in
sociology, with seminal works such as the ones from Kluegel
(\citeproc{ref-kluegel_social_1995a}{1995};
\citeproc{ref-kluegel_beliefs_1986}{1986}) and Lane
(\citeproc{ref-lane_political_1962}{1962};
\citeproc{ref-lane_market_1986}{1986}). Lane's main aim was to ``help
explain the tenacious hold of market capitalism on the public mind''
(\citeproc{ref-lane_market_1986}{1986, p. 384}), proposing the
distinction between market and political justice. Whereas political
justice refer to government actions aligned with justice principles such
equality or need, market justice involves the allocation linked to
earned desserts, particularly through individual contributions based on
as effort and ability. Sociological survey research in this area has
considered questions related to the extent to which it is considered
just to get better access to social services (such as pensions, health
and education) based on payment capacity
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). Although
these services could be considered basic rights of citizens, evidence to
date suggests that a significant portion of the population justifies
inequality in these areas (\citeproc{ref-koos_moral_2019}{Koos \&
Sachweh, 2019}) and that such preferences vary according to objective
socioeconomic factors and the contextual characteristics of countries
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In the present paper, we
argue that there are two other relevant factors to take into account
when analyzing differences between individuals and groups to this
regard: the perception of economic inequality and the perception of
meritocracy.

The perception of economic inequality can be understood as an
individual's subjective assessment of how resources are allocated among
members of a given society (\citeproc{ref-akyelken_urban_2020}{Akyelken,
2020}). Regardless of their measurement, various studies have shown that
this perception often underestimates the gap between the rich and the
poor, revealing inconsistencies between objective and subjective aspects
of inequality (\citeproc{ref-trump_income_2018}{Trump, 2018}). Moreover,
there is consistent evidence that perceived inequality, even under
control of objective inequality, could have implications for attitudes
toward the distribution of goods and services
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-schroder_income_2017}{Schröder, 2017}), redistributive
preferences, justification of inequality, and legitimacy of the economic
system (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_attitudes_2020}{García-Sánchez et al.,
2020}, \citeproc{ref-garcia-sanchez_perceptions_2018}{2018}). From this
perspective, in this paper we argue that perceptions of economic
inequality may play a relevant aspect for understanding market justice
preferences, as a lower perceived inequality could reinforce the idea
that the differences in access to social services could be justified ---
a link that becomes difficult to hold when high inequality is perceived.

Along with perception of inequality, a second focus of this paper is the
perception of meritocracy. Meritocracy posits that inequality can be
legitimized through distributive criteria such as effort and talent
(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
\citeproc{ref-young_rise_1962}{Young, 1962}). Previous studies have
demonstrated that people with stronger meritocratic beliefs tend to
perceive less inequality as they attribute economic differences to
individual achievements (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
\citeproc{ref-wilson_role_2003}{Wilson, 2003}) and also to justify more
inequality as these beliefs are associated with attitudes that
legitimize status differences
(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In a context
as Chile, where the distribution of goods and services is predominantly
governed by market logics strongly introduced during the military
dictatorship (1973-1989) (\citeproc{ref-boccardo_30_2020}{Boccardo,
2020}; \citeproc{ref-madariaga_three_2020}{Madariaga, 2020}),
meritocratic perceptions have the potential to affect the acceptance of
social inequalities. Therefore, incorporating the perception of
meritocracy in this literature allows for an understanding of how
attitudes toward inequality are shaped not only by the perception of
economic gaps, but also by beliefs about how these gaps are justified.
Furthermore, these beliefs are often consolidated from an early age,
reinforced by institutions promoting values such as effort and
individual skills to climb socially
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian, 2014}).

At this point we can propose the following research questions: To what
extent the preferences for market justice are associated with
perceptions of inequality and meritocracy? How stable are these
associations (within subjects) along time? The primary objective of this
study is to analyze the interplay between perceptions of inequality and
meritocracy and their joint influence on preferences for market justice
along time in Chile using the panel survey ELSOC (Social Chilean
Longitudinal Study, 2016-2023). The chilean case appears as a particular
case for studying market justice preferences given their story of high
inequality and strong free-market policies introduced in the
dictatorship period (\citeproc{ref-flores_top_2020}{Flores et al.,
2020}; \citeproc{ref-madariaga_three_2020}{Madariaga, 2020}).
Additionally, this analysis seeks to explore how political and social
contingencies---such as the 2019 and 2022 social movements---might have
moderated these relationships by prompting more critical reflection on
the commodification of social services
(\citeproc{ref-somma_no_2021}{Somma et al., 2021}). Recognizing the
temporal dimension in shaping market justice preferences is essential,
given that historical and contextual factors that could challenge or
reaffirm the way in which individuals persceive inequality and
meritocracy, and consequantely --- following our argument --- the
preferences for market justice.

@@

Conceptually, \emph{market justice} has been discussed in the literature
as a normative principle that legitimates the distribution of economic
rewards based on individual merit. It is possible to trace the concept
to the understanding of Lane (\citeproc{ref-lane_market_1986}{1986}),
who makes a contrast between market justice and political justice. The
author defines market justice as a system of ``earned deserts'', whereby
individuals are seen as deserving of a determined distributive outcome
due to their effort and skills. In contrast, political justice
emphasizes principles of equality and need, which are often represented
by the welfare state action through social policies. An important remark
is that the principles of market justice rely on the assumption that
markets are neutral and self-regulating arenas, where individuals are
treated fairly because they face the same formal rules of engagement
(\citeproc{ref-lane_market_1986}{Robert E. Lane, 1986}). Consequently,
the legitimacy of market justice stems from the assumption that
inequalities are not only inevitable but fair---so long as the rules are
transparent and opportunities are open. In this way, market justice
provides a moral justification for inequality by framing it as a
necessary outcome of individual responsibility
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

@@

The study of market justice preferences has increasingly focused on how
they are shaped by individuals' socioeconomic position, normative
beliefs, and the institutional context in which they are embedded.
Across the literature, there is empirical evidence suggesting that
individuals in more advantaged socioeconomic positions---those with
higher occupational class, income, and education---are more likely to
support market justice principles (\citeproc{ref-koos_moral_2019}{Koos
\& Sachweh, 2019}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). This tendency
reflects not only material self-interest but also a broader moral
economy, in which winners of the market system internalize
justifications for the status quo. At the same time, political ideology
also plays a role ---such as economically conservative values --- where
more right-wing individuals show higher support for meritocracy and more
skepticism towards redistribution. This is particularly salient in
countries with more restricted public provision of social services. For
example, Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) in Chile and, Lee and
Stacey (\citeproc{ref-lee_fairness_2023}{2023}) in Australia show that
right-wing individuals are more supportive of market-based distribution
of welfare. Beyond individual characteristics, country-level
institutions also play a central role. In liberal welfare regimes like
those of the United States or the United Kingdom, market justice
preferences are more widespread, while in coordinated or
social-democratic regimes---such as Sweden or Germany---citizens are
generally more critical of market-based inequalities
(\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
that market justice is also shaped contextually, which is represented by
the development of welfare institutions, in line with the
policy-feedbacks literature
(\citeproc{ref-campbell_policy_2012}{Campbell, 2012};
\citeproc{ref-pierson_when_1993}{Pierson, 1993}).

@@

Perceptions of inequality have been associated with a range of
distributive attitudes (\citeproc{ref-castillo_perception_2022}{Castillo
et al., 2022}; \citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez
et al., 2019}; \citeproc{ref-mccall_exposure_2017}{McCall et al., 2017};
\citeproc{ref-trump_income_2018}{Trump, 2018}). Research indicates that
perceptions of inequality influence redistributive preferences to a
greater extent than objective or actual economic inequality
(\citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \& Treisman,
2018}). Others have shown that the perception of inequality, through the
perceived wage gap, is related to the justification of greater
inequality (\citeproc{ref-Castillo2011}{Castillo, 2011}). In that sense,
lower perceived inequality can reinforce support for market-based
distributive arrangements by suggesting that the system is fair and that
outcomes reflect effort and ability (\citeproc{ref-kuhn_eye_2011}{Kuhn,
2011}). In contrast, when inequality is perceived as excessive or
structurally determined, individuals could question the legitimacy of
market justice and become more supportive of redistributive policies
(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}).

Perceptions of inequality refer to individuals' subjective evaluations
of the extent, causes, and consequences of income and wealth disparities
(\citeproc{ref-akyelken_urban_2020}{Akyelken, 2020}). Unlike objective
measures such as the Gini index, perceived inequality captures how
individuals make sense of distributive hierarchies in their everyday
lives, shaped by reference groups, social comparisons, and information
environments (\citeproc{ref-garcia-castro_perceiving_2020}{García-Castro
et al., 2020}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
Treisman, 2018}; \citeproc{ref-mijs_stratified_2016}{Mijs, 2016}).
Scholars have proposed multiple dimensions of perceived inequality,
including its magnitude (how significant are the gaps), vertical
structure (between which groups), the trend over time (increasing or
decreasing), and legitimacy (whether it is just or not)
(\citeproc{ref-engelhardt_what_2018}{Engelhardt \& Wagener, 2018};
\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
2019}). These dimensions encompass both cognitive and normative aspects
of perceptions of inequality and can vary across societies and social
groups, depending on exposure, ideology, and personal experience
(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
2018}).

@@

In addition to influencing individual attitudes toward inequality,
meritocratic perceptions can contribute to social division and the
stigmatization of disadvantaged groups. Recent research has demonstrated
that exposure to meritocratic narratives can reinforce the belief that
poverty results from individual failings rather than systemic
conditions, reducing support for redistributive measures and increasing
the stigmatization of the poor (\citeproc{ref-hoyt_mindsets_2023}{Hoyt
et al., 2023}). This reinforces negative stereotypes and reduces empathy
toward individuals from lower socioeconomic backgrounds. Moreover,
Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argues
that meritocratic narratives can serve as feedback mechanisms that shape
public opinion and well-being by framing individuals' understanding of
welfare outcomes as deserved or undeserved within existing institutional
structures. This psychological mechanism highlights the normative power
of meritocracy in stabilizing unequal systems by shaping political
attitudes and personal perceptions of success and failure.

\subsection{The Chilean context}\label{the-chilean-context}

Chile offers an interesting case for examining how public attitudes
toward the distribution of social services evolve amid declining poverty
and persistently high income inequality within a residual welfare
framework (\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). Following
the neoliberal reforms of the 1980s, the Chilean welfare state has
become increasingly reliant on private provision, with essential
services often commodified and largely reserved for those with
sufficient purchasing power (\citeproc{ref-boccardo_30_2020}{Boccardo,
2020}; \citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). This model
has favored the affluent, while lower-income groups are left dependent
on limited and underfunded public alternatives. Research on social
stratification characterizes Chile as a society where mobility into
intermediate classes is feasible, but access to elite strata remains
largely restricted
(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}; \citeproc{ref-torche_intergenerational_2014a}{Torche, 2014}).
While some indicators---such as education and occupation---suggest a
relatively fluid class structure, income mobility remains constrained
(\citeproc{ref-espinoza_movilidad_2014}{Espinoza \& Núñez, 2014}).

Despite steady economic expansion, Chile continues to rank among the
most unequal nations in the OECD, marked by a high Gini coefficient and
substantial wealth concentration at the top
(\citeproc{ref-flores_top_2020}{Flores et al., 2020}). In this context,
Chile experienced a wave of mass protests from October 2019 to March
2020. Initially led by high school students, the demonstrations quickly
gained momentum as broader segments of society mobilized around demands
for more equality of opportunities, particularly regarding precarious
access to welfare services. The political system interpreted this unrest
as a call to reform existing political institutions, with a special
focus on equalizing opportunities and providing more egalitarian access
to welfare by reducing reliance on out-of-pocket mechanisms
(\citeproc{ref-somma_no_2021}{Somma et al., 2021}).

Concerning public opinion, the ELSOC survey shows that Chile has a low
level of agreement with the idea that it is fair for those with higher
incomes to buy better welfare in the areas of education, healthcare, and
pensions, with average support of 8.5\% between 2016 and 2023. At the
same time, around 55\% and 47\% of the public between 2016 and 2023
\emph{disagreed} that effort and talent are rewarded accordingly. This
coincides with 90\% of citizens \emph{agreeing} that income differences
in Chile are \emph{too large}
(\citeproc{ref-castillo_changes_2025}{Castillo et al., 2025}).

In sum, Chile represents an illustrative case of a particular form of
\emph{moral economy} (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh,
2019}), in which national socioeconomic characteristics, institutional
arrangements, and public attitudes coexist. This suggests that the
country combines persistently high levels of economic and social
inequality with a strong institutionalization of market-based mechanisms
within the welfare system
(\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). At the same
time, perceptions of inequality are high, and there is widespread public
criticism of the commodification of social services
(\citeproc{ref-somma_no_2021}{Somma et al., 2021}). For the above
reasons, we argue that the Chilean case represents a relevant context to
study the role of perceived income inequality and meritocracy in shaping
public preferences regarding the commodification of welfare services,
particularly from the perspective of individual change.

@@

\(H1_{a} \; (\text{Between-person}):\) Individuals who perceive higher
levels of economic inequality will report lower support for market
justice preferences, on average, compared to those who perceive lower
levels of inequality.
